---
title: "Project Analysis"
author: "Ashish Chaudhary"
output:
  pdf_document:
    latex_engine: xelatex
  html_notebook: default
---

```{r}
library(tidyverse)
library(lubridate)#date/time handling
library(vader)#sentiment analysis
library(tm)#text preprocessing
library(stringr)# For string manipulation

covid_tweets = read.csv("covidvaccine.csv", stringsAsFactors = FALSE, na.strings = c("","NA","NaN"))
```

#EDA
```{r}
glimpse(covid_tweets)
names(covid_tweets)
head(covid_tweets)
colSums(is.na(covid_tweets))
```

# Handling Missing data
```{r}
#Remove rows with missing tweets, user_name (including unverified ones)
covid_tweets_clean <- covid_tweets %>% drop_na(text,user_name,user_verified)

# Dropping insignificant columns
covid_tweets_clean <- covid_tweets_clean %>% select(-user_created, -date) # 90% values missing

# Retweets may not represent original sentiment from that user, proceed with original tweets only.
covid_tweets_clean <- covid_tweets_clean %>%select(-"is_retweet")

covid_tweets_clean$user_friends <- as.numeric(covid_tweets_clean$user_friends)
covid_tweets_clean$user_favourites <- as.numeric(covid_tweets_clean$user_favourites)

#imputing with median
covid_tweets_clean <- covid_tweets_clean %>%
  mutate(
    user_followers = ifelse(is.na(user_followers), median(user_followers, na.rm = TRUE), user_followers),
    user_friends = ifelse(is.na(user_friends), median(user_friends, na.rm = TRUE), user_friends),
    user_favourites = ifelse(is.na(user_favourites), median(user_favourites, na.rm = TRUE), user_favourites),
    follower_friend_ratio= user_followers / (user_friends + 1) # follower-to-friend ratio gives better idea about engagement of user.
  )

glimpse(covid_tweets_clean)
```

```{r}
#missing values in each column
colSums(is.na(covid_tweets_clean))

# removing duplicate tweets
sum(duplicated(covid_tweets_clean$text))
covid_tweets_clean <- covid_tweets_clean %>% distinct(text, .keep_all = TRUE)
glimpse(covid_tweets_clean)
```
In total 417 records removed.

# Text Based Features
```{r}
# 1. Calculate text length (character count)
covid_tweets_clean$text_length <- nchar(covid_tweets_clean$text)

# 2. Calculate word count
covid_tweets_clean$word_count <- str_count(covid_tweets_clean$text, "\\S+")

# 3. Count number of hashtags
covid_tweets_clean$hashtag_count <- str_count(covid_tweets_clean$hashtags, ",") + 
  ifelse(covid_tweets_clean$hashtags == "" | covid_tweets_clean$hashtags == "[]", 
         0, 1)

# 4. Check if tweet contains URLs
covid_tweets_clean$has_url <- grepl("http", covid_tweets_clean$text, ignore.case = TRUE)

# 5. Check if tweet contains mentions (@)
covid_tweets_clean$has_mention <- grepl("@", covid_tweets_clean$text)

# 6. Count exclamation marks (indicates emotion/emphasis)
covid_tweets_clean$exclamation_count <- str_count(covid_tweets_clean$text, "!")

# 7. Count question marks
covid_tweets_clean$question_count <- str_count(covid_tweets_clean$text, "\\?")
names(covid_tweets_clean)
```


# Sentiment Analysis Using VADER 
```{r}
library(sentimentr)

# Step 1: Pre-parse tweets into sentences
sentences <- get_sentences(covid_tweets_clean$text)

# Step 2: Sentiment analysis (fast)
sentiment_scores <- sentiment_by(sentences)

# Step 3: Add to dataset
covid_tweets_clean$sentiment_compound <- sentiment_scores$ave_sentiment

# Optional sentiment categories like VADER
covid_tweets_clean$sentiment_category <- cut(covid_tweets_clean$sentiment_compound,
  breaks = c(-Inf, -0.05, 0.05, Inf),
  labels = c("Negative", "Neutral", "Positive")
)

```


# Analysis of Sentiment using Frequentist and Bayesian Linear regression
```{r}
library(lmtest); library(corrplot);
library(BayesFactor); library(performance);
library(MCMCpack); library(bayestestR);
library(easystats); library(coda)
```

Positive correlation → higher value → more positive sentiment

Negative correlation → higher value → more negative sentiment

Magnitude: 0.00–0.10 = very weak, 0.10–0.30 = weak, 0.30+ = moderate+



```{r}
model_data <- covid_tweets_clean %>%
  mutate(
    log_followers = log10(user_followers + 1),
    log_friends = log10(user_friends + 1)
  ) %>%
  drop_na()
```


```{r}
#important to have numeric coulmns for corr
numeric_vars <- model_data %>% select_if(is.numeric)
cor_matrix <- cor(numeric_vars, use = "complete.obs")
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.7, tl.col = "black")
cor_matrix[,"sentiment_compound"]
```
Correlations are very small, meaning no variable strongly predicts sentiment alone.
Hashtags and questions lean slightly negative, exclamation marks lean slightly positive.
User level data (followers, friends) has almost no meaningful correlation with sentiment, hence could be removed for further analysis

# FREQUENTIST LINEAR REGRESSION ANALYSIS

Null hyp : No relationship between any predictor and sentiment
Alt hyp: At least one predictor significantly affects sentiment

```{r}
model_formula <- sentiment_compound ~ text_length + word_count + hashtag_count +has_url +
                has_mention + exclamation_count + question_count + log_followers + log_friends +
                follower_friend_ratio

freq_model <- lm(model_formula, data = model_data)
summary(freq_model)
```

The overall model is significant (F(10, 259947) = 469.5, p<0.05). This confirms that at least one predictor affects sentiment.
However, R2 is 0.0177 indicates the model explains only 1.84% of variance, showing weak practical significance despite statistical significance

hashtag_count(B=-0.00665, p<0.05): More hashtags, slightly more negative sentiment.
exclamation_count(B=0.0139, p<0.05): Tweets with more ! are more positive.
question_count(B=-0.0172, p<0.05): Questions express uncertainty or concern,more negative sentiment.
has_url(B=0.0239, p<0.05):Tweets with URLs are marginally more positive. Similary has_mention(B=0.0096, p<0.05).

follower_friend_ratio(B approx. 0, p<0.05):statistically significant due to large N, but practically meaningless effect.

text_length (B=0.000146, p<0.05)
word_count (B=-0.000728, p<0.05)
Both are significant but the effect sizes are almost zero, offering negligible practical value.

We reject the null hypothesis statistically, the model detects multiple predictors affecting sentiment.

```{r}
check_collinearity(freq_model)# VIF and multicollinearity
check_residuals(freq_model)#autocorrelation + distribution of residuals
check_outliers(freq_model)
```

Since VIF of all variables are less than 5, we can say them largely independent and do not exhibit harmful multicollinearity, except text_length and word_count with 14.14 and 11.93 respectively, so we can toss them out from analysis.

The model shows mild heteroscedasticity(non-uniformity of variance), likely due to large sample size.

No Influential Outliers.

# BAYESIAN LINEAR REGRESSION ANALYSIS
```{r}
model_data <- model_data %>%
  mutate(
    has_url = as.factor(has_url),
    has_mention = as.factor(has_mention)
  ) %>% filter(if_all(where(is.numeric), is.finite))
bf_model <- lmBF(model_formula, data = model_data)
summary(bf_model)
```

The Bayes factor for the full model is 8.906276e+1053, which is huge, indicating overwhelming evidence in favor of the model with these predictor(text_length + word_count + hashtag_count + has_url + has_mention + exclamation_count + question_count + log_followers + log_friends +     follower_friend_ratio) i.e. in favor of alt hypothesis.

```{r}
lmout_mcmc= lmBF(sentiment_compound ~ word_count + hashtag_count + has_url + has_mention +
            exclamation_count + question_count + log_followers + log_friends +
            follower_friend_ratio, data = model_data, posterior = TRUE, iterations = 10000)

summary(lmout_mcmc)
mcmc_chain <- as.mcmc(lmout_mcmc)
traceplot(mcmc_chain)
describe_posterior(lmout_mcmc)
```
The trace Plots looks good.

The Bayesian model indicates that most predictors have small but meaningful effects on sentiment_compound. Positive associations are observed for exclamation_count, log_followers, and has_mention-TRUE, while negative associations are seen for word_count, hashtag_count, question_count, and log_friends. 

For some predictors (word_count and follower_friend_ratio), the 95% CI are very narrow include 0 or super close to 0, suggesting that their effects are weak or potentially negligible. Overall, the model captures subtle but consistent relationships, though practical effect sizes are small.

All posterior distributions are narrow with high probability density, indicating stable estimates, and nearly all parameters lie outside the ROPE of [-0.1, 0.1], confirming meaningful effects. Overall, the model captures subtle but consistent relationships between tweet/user features and sentiment, though effect sizes are small.



################################################ Conclusion #########################################################

>> Null hyp : No relationship between any predictor and sentiment
   Alt hyp: At least one predictor significantly affects sentiment

>> Overall Model Significance: 
Both analyses show the model is statistically significant (rejecting null hypothesis), indicating at least one predictor affects sentiment. Frequentist R2 is approx 0.018 shows low explanatory power.

>> Positive Predictors:
exclamation_count and log_followers are consistently associated with more positive sentiment.

>> Negative Predictors: 
word_count, hashtag_count, and log_friends show negative associations with sentiment, though some intervals include 0 in Bayesian analysis, indicating uncertainty.

>> Non-Significant:
Features like text_length and follower_friend_ratio have minimal or uncertain effects. CIs contain 0.



